### 2015-06-06 chapters/ipa_meets_unicode.tex
updating IPA diacritics pitfall

It is not just the Unicode Standard that offers multiple options for encoding
the IPA. Even the IPA specification itself offers some flexibility in how
transcriptions have to be encoded. There are a few cases in which the IPA
explicitly allows for different options of transcribing the same phonetic
content. This is understandable from a transcriber's point of view, but it is
not acceptable for interoperability between resources written in IPA. We
consider it crucial to distinguish between ``lax'' IPA, for which it is
sufficient that any phonetically-trained reader is able to understand the
transcription, and ``strict'' IPA, which should be standardized on a single
unique encoding for each sound, so search will work across resources.

The first case in which the IPA allows for different encodings is the question
of how to transcribe tone. There is an old tradition to use diacritics on vowels
to mark different tone levels, e.g. <ȅèée̋>.\footnote{There are at least two
different Unicode homoglyphs for the low and high level tones, namely <\ ̀~>
\textsc{combining grave tone mark} at \unif{0340} vs.\@ <\ ̀~> \textsc{combining
grave accent} at \unif{0300} for low tone, and <\ ́~> \textsc{combining acute tone
mark} at \unif{0341} vs.\@ <\ ́~> \textsc{combining acute accent} at \unif{0301}
for high tone.} The IPA also proposes a less widespread used, but more
consistent option of tone letters, e.g. <˥˦˧˨˩>. Tone letters in the IPA have
five different levels, and sequences of these letters can be used to indicate
contours. Well-designed fonts will even merge a sequence of tone letters into a
contour. For example, compare the font Linux Libertine, which does not merge
tone letters <{\fontspec{LinLibertineO}˥˨˧˩}>, with the font CharisSIL, which
merges this sequence of four tone letters into a single contour <\charis{˥˨˧˩}>.
For strict IPA encoding we propose to standardize on tone letters.

@@

Reversely, <ç> \textsc{latin small letter c with cedilla} at \uni{00E7} will be
decomposed into <c> \textsc{latin small letter c} at \uni{0063} with 
<\ \ {\large  ̧}> \textsc{combining cedilla} at \uni{0327}, 
also if such a decompotion is not intended, because it is meaningless in IPA.

\section{Pitfall: Different notions of diacritics}
\label{pitfall-the-ipa-notion-of-diacritics-is-not-the-same-as-the-unicode-standards-notion-of-diacritics}

Another pitfall is diacritics. The problem is that the meaning of the term
\textsc{diacritics} as used by the IPA is not the same as it used in the Unicode
Standard. Specifically, diacritics in the IPA-sense are either so-called
\textsc{Combining Diacritical Marks} or \textsc{Spacing Modifier Letters} in the
Unicode Standard. Crucially, Combining Diacritical Marks are by definition
combined with the character before them (to form so-called default grapheme
clusters, see Section 3). In contrast, Spacing Modifier Letters are by
definition \emph{not} combined into grapheme clusters with the preceding
character, but simply treated as separate letters. In the context of the IPA,
the following IPA-diacritics are actually Spacing Modifier Letters in the
Unicode Standard:

\begin{itemize}
	\item Length marks, namely <ː> \textsc{modifier letter triangular colon} at \uni{02D0} and <ˑ> \textsc{modifier letter half triangular colon} at \uni{02D1}. 
	\item Tone letters, like <˥> \textsc{modifier letter extra-high tone bar} at \uni{02E5}, and others like this. 
	\item Superscript letters, like <ʰ> \textsc{modifier letter small h} at \uni{02B0} or <ˤ> \textsc{modifier letter small reversed glottal stop} at \uni{02E4}, and others like this. 
	\item <˞> \textsc{modifier letter rhotic hook} at \uni{02DE}. 
\end{itemize}

Although linguists might expect these characters to belong together with the
character in front of them, at least for <ʰ> \textsc{modifier letter small h} at
\uni{02B0} the Unicode Consortium's decision to treat it as a separate character
is also linguistically correct, because according to the IPA it can be used both
for aspiration (more precisely post-aspiration following the base character) and
pre-aspiration (preceding the base character). Note that there is a mechanism in
Unicode to force separate characters to be combined (namely by using the
\textsc{zero width joiner} at \uni{200D}), but this seems to be a rather
impractical, and probably not enforceable solution to us.

>>>>

It is not just the Unicode Standard that offers multiple options for encoding
the IPA. Even the IPA specification itself offers some flexibility in how
transcriptions have to be encoded. There are a few cases in which the IPA
explicitly allows for different options of transcribing the same phonetic
content. This is understandable from a transcriber's point of view, but it is
not acceptable for interoperability between resources written in IPA. We
consider it crucial to distinguish between ``lax'' IPA, for which it is
sufficient that any phonetically-trained reader is able to understand the
transcription, and ``strict'' IPA, which should be standardized on a single
unique encoding for each sound, so search will work across resources. We are 
aware of the following double options in the IPA, which will be discussed in 
turn below:

\begin{itemize}
  \item The marking of tone
  \item The marking of <g>
  \item The diacritic for voiceless
  \item The tie bar to indicate a close bond between sounds
\end{itemize}

\

\noindent The first case in which the IPA allows for different encodings is the question
of how to transcribe tone. There is an old tradition to use diacritics on vowels
to mark different tone levels, e.g. <ȅèée̋>.\footnote{There are at least two
different Unicode homoglyphs for the low and high level tones, namely <\ ̀~>
\textsc{combining grave tone mark} at \unif{0340} vs.\@ <\ ̀~> \textsc{combining
grave accent} at \unif{0300} for low tone, and <\ ́~> \textsc{combining acute tone
mark} at \unif{0341} vs.\@ <\ ́~> \textsc{combining acute accent} at \unif{0301}
for high tone.} The IPA also proposes a less widespread used, but more
consistent option of tone letters, e.g. <˥˦˧˨˩>. Tone letters in the IPA have
five different levels, and sequences of these letters can be used to indicate
contours. Well-designed fonts will even merge a sequence of tone letters into a
contour. For example, compare the font Linux Libertine, which does not merge
tone letters <{\fontspec{LinLibertineO}˥˨˧˩}>, with the font CharisSIL, which
merges this sequence of four tone letters into a single contour <\charis{˥˨˧˩}>.
For strict IPA encoding we propose to standardize on tone letters.

@@

Reversely, <ç> \textsc{latin small letter c with cedilla} at \uni{00E7} will be
decomposed into <c> \textsc{latin small letter c} at \uni{0063} with 
<\ \ {\large  ̧}> \textsc{combining cedilla} at \uni{0327}, 
also if such a decomposition is not intended, because it is meaningless in IPA.

\section{Pitfall: Different notions of diacritics}
\label{pitfall-different-notions-of-diacritics}

Another pitfall relates the concept of what are diacritics. The problem is that
the meaning of the term diacritics as used by the IPA is not the same as is used
in the Unicode Standard. Specifically, diacritics in the IPA-sense are either
so-called \textsc{combining diacritical marks} or \textsc{spacing modifier
letters} in the Unicode Standard. Crucially, Combining Diacritical Marks are by
definition combined with the character before them (to form so-called default
grapheme clusters, see Section~\ref{the-unicode-approach}). In contrast, Spacing
Modifier Letters are by definition \emph{not} combined into grapheme clusters
with the preceding character, but simply treated as separate letters. In the
context of the IPA, the following IPA-diacritics are actually Spacing Modifier
Letters in the Unicode Standard:

\begin{itemize}
	\item[] Length marks, namely 
	\begin{itemize}
	  \item[] <ː> \textsc{modifier letter triangular colon} at \uni{02D0}
	  \item[] <ˑ> \textsc{modifier letter half triangular colon} at \uni{02D1}
	\end{itemize}

	\item[] Tone letters, like 
	\begin{itemize} 
	  \item[] <˥> \textsc{modifier letter extra-high tone bar} at \uni{02E5}
	  \item[] <˧> \textsc{modifier letter mid tone bar} at \uni{02E7}
	  \item[] and others like this
	\end{itemize}

	\item[] Superscript letters, like 
	\begin{itemize}
	  \item[] <ʰ> \textsc{modifier letter small h} at \uni{02B0}
	  \item[] <ˤ> \textsc{modifier letter small reversed glottal stop} at \uni{02E4}
	  \item[] and many more like this
	\end{itemize}

	\item[] the rhotic hook <˞> \textsc{modifier letter rhotic hook} at \uni{02DE}
\end{itemize}

Although linguists might expect these characters to belong together with the
character in front of them, at least for <ʰ> \textsc{modifier letter small h} at
\uni{02B0} the Unicode Consortium's decision to treat it as a separate character
is also linguistically correct, because according to the IPA it can be used both
for aspiration (more precisely post-aspiration following the base character) and
pre-aspiration (preceding the base character). Note that there exists a mechanism in
Unicode to force separate characters to be combined (namely by using the
\textsc{zero width joiner} at \uni{200D}), but this seems to be a rather
impractical, and probably not an enforceable solution to us.
